---
title: "Simulated Spotify Wrapped"
author: Paityn Richardson
subtitle: December 9, 2024
format:
  revealjs:
    scrollable: true
    slide-number: true
    show-slide-number: all
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

# Step 1

## Generate Listening History

-   want to simulate a YEAR worth of listening history

-   60,000 / 3.5 = 17,100

```{r}
my_spotify_library <- read.csv("~/paitynleigh.github.io/MyspotifyLibrary.csv")

colnames(my_spotify_library)
```

::: aside
Tool "Tune My Music" Used to collect playlist data https://www.tunemymusic.com/transfer
:::

------------------------------------------------------------------------

```{r}
library(tidyverse)
library(gt)

simulated_year <- function(my_spotify_library, n = 17100) {
  my_spotify_library|>
    slice_sample(n = n, replace = TRUE)
}

```

# Step 2

## Data Conclusions

-   Recreating Spotify Wrapped

-   Top 5 Songs, Artists, and Albums

```{r}
top_5 <- function(data, colname){
  data |>
    count({{colname}}, sort = TRUE)|>
    head(5)
}

```

-   Can use this same function for songs, artists, and albums

## Now, Lets see 1 iteration of Simulated Listening History

```{r}
find_top_5 <- function(simulated_year) {
  top_songs <- top_5(simulated_year, Track.name)
  top_artists <- top_5(simulated_year, Artist.name)
  top_albums <- top_5(simulated_year, Album)
  
  # labels each table 
  list(songs = top_songs, artists = top_artists, albums = top_albums)
}
find_top_5(simulated_year(my_spotify_library))

```

# Step 3

## Mapping

-   Lets create multiple simulated years of listening

```{r}
iters <- 1000

results <- map(1:iters, ~ {
  simulated_year <- simulated_year(my_spotify_library) 
  find_top_5(simulated_year) 
})

```

-   What questions can mapping help us answer?

## Probability

-   How likely is it that a specific artist will be in my top 5? Top 1??

```{r}

check_artist <- function(top_artists, artist) {
  in_top_5 <- top_artists|>
    filter(Artist.name == artist) |>
    # Will return Boolean for this statement
    nrow() > 0  
  
  is_top_artist <- top_artists |>
    pull(Artist.name) |>
    head(1) == artist 
  
  list(
    in_top_5 = in_top_5,
    is_top_artist = is_top_artist
  )
}

```

## cont.

```{r}
artist <- "TV Girl"

artist_occurrences <- map(results, function(results) {

  check_artist(pluck(results, "artists"), artist)
})

data <- map_dfr(artist_occurrences, ~ tibble(
  in_top_5 = pluck(.x, "in_top_5"),
  is_top_artist = pluck(.x, "is_top_artist")
))

data|>
  head(1)
```

## Conclusions

```{r}
prob_top_5 <- data |>
  select(in_top_5) |>
  summarise(mean = mean(in_top_5))|>
  pull(mean)

prob_top_1 <- data |>
  select(is_top_artist) |>
  summarise(mean = mean(is_top_artist))|>
  pull(mean)

prob_top_5
prob_top_1
```
