---
title: "WAI Database Study"
description: |
  Analysis of Sound Absorbance Data from WAI Database
author: Paityn Richardson
date: November 26, 2018
---

```{r}
library(RMariaDB)
library(tidyverse)
con_wai <- dbConnect(
  MariaDB(), host = "scidb.smith.edu",
  user = "waiuser", password = "smith_waiDB", 
  dbname = "wai"
)
Measurements <- tbl(con_wai, "Measurements")
PI_Info <- tbl(con_wai, "PI_Info")
Subjects <- tbl(con_wai, "Subjects")

# collect(Measurements)
```

In this project I plan to use the data from the [Wideband Acoustic Immittance (WAI) Database](https://www.science.smith.edu/wai-database/) hosted by Smith College. With this data I plan to replicate the figure from Voss (2020) titled "Mean Absorbance from Each Publication in the WAI Database". I started of by arranging the data using a SQL query.

```{sql}
#| connection: con_wai
#| output.var: "new_table"
    SELECT 
        m.Frequency,
        m.Identifier,
        m.Instrument,
        AVG(m.Absorbance) AS Mean_Absorption, 
        CONCAT(pi.AuthorsShortList, ' (', pi.Year, ')', ' N=', COUNT(DISTINCT CONCAT(m.SubjectNumber, '-', m.Ear)),'; ', m.Instrument) AS Label
    FROM 
        Measurements m
    JOIN 
        PI_Info pi
        ON m.Identifier = pi.Identifier
    WHERE 
        m.Identifier IN ('Abur_2014', 'Feeney_2017', 'Groon_2015', 'Lewis_2015', 'Lui_2008', 'Rosowski_2012', 'Shahnaz_2006', 'Shaver_2013', 'Sun_2016', 'Voss_1994', 'Voss_2010', 'Werner_2010')
    GROUP BY 
        m.Frequency, 
        m.Identifier, 
        m.Instrument, 
        pi.Year, 
        pi.AuthorsShortList
    ORDER BY 
        m.Frequency, 
        m.Identifier, 
        m.Instrument


```

In this query, I grouped the data by Frequency, Identifier, Instrument, Year, and AuthorShortList. In order to get the variable "AuthorShortList" I joined the PI_Info table and the Measuements table on SubjectNumber. I used Concat to make a label for each of the studies with the number of unique ears, the instrument used, and the year and author of the study.

```{r}
library(ggplot2)


ggplot(new_table, aes(x = Frequency, y = Mean_Absorption, color = Label)) +
  geom_line(size = .7, na.rm = TRUE) +  
  scale_x_continuous(
    name = "Frequency (Hz)",
    breaks = c(0, 2000, 3000, 3500, 3850, 6000, 7000, 7500, 8000),
    labels = c("200", "400", "600", "800", "1000", "2000", "4000", "6000", "8000")
  ) +
  scale_y_continuous(
    name = "Mean Absorbance",
    limits = c(0, 1), 
    breaks = seq(0, 1, by = 0.1)  
  ) +
  labs(
    title = "Mean Absorbance From Each Publication in the WAI Database",  
    color = NULL 
  ) +
  theme_minimal() + 
  theme(
    axis.text = element_text(size = 5),
    axis.title = element_text(size = 7),
    plot.title = element_text(size = 9, face = "bold", hjust = .5),
    legend.text = element_text(size = 5),
    legend.position = c(0.28, 0.9),
    legend.key.size = unit(0.1, "cm"),
    legend.spacing = unit(0.01, "cm"),
    legend.background = element_rect(color = "black", size = .2, fill = "white"),
    aspect.ratio = 1
  ) +
  coord_cartesian(xlim = c(200, 8000))



```

This table shows the mean absorbance of frequencies ranging from 200 HZ to 8000 HZ in 12 different studies in the WAI database. As seen in the table, absorbance, on average, increases as frequency increases up until around 1000 HZ, when it drops off and begins to decrease again.

```{sql}
#| connection: con_wai
#| output.var: "new_table4"
SELECT 
    m.Frequency, 
    s.Sex, 
    AVG(m.Absorbance) AS Avg_Absorbance
FROM 
    Measurements m
JOIN 
    Subjects s
ON 
    m.SubjectNumber = s.SubjectNumber
WHERE 
    m.Identifier = 'Aithal_2015'
    AND s.Sex != 'Unknown'
GROUP BY 
    m.Frequency, s.Sex
ORDER BY 
    m.Frequency, s.Sex;

```

Now with this SQL query, I am doing a similar thing to the last, except now I am also grouping by "sex" which required be to join the table "Subjects" also in the WAI database. I also chose only 1 specific study to measure on this table, Aithal (2014).

```{r}
dbDisconnect(con_wai, shutdown = TRUE)
```

```{r}
library(ggplot2)

ggplot(new_table4, aes(x = Frequency, y = Avg_Absorbance, color = Sex)) +
  geom_line(size = 0.7, na.rm = TRUE) +
  scale_x_continuous(
    name = "Frequency (Hz)",
    breaks = c(200, 400, 600, 800, 1000, 2000, 4000, 6000, 8000),
    labels = c("200", "400", "600", "800", "1000", "2000", "4000", "6000", "8000")
  ) +
  scale_y_continuous(
    name = "Average Absorbance",
    limits = c(0, 1), 
    breaks = seq(0, 1, by = 0.1)
  ) +
  labs(
    title = "Average Absorbance by Frequency and Gender",
    color = NULL 
  ) +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 8),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 8),
    legend.position = c(0.1, 0.9),
    legend.box.background = element_rect(color = "black", fill = "white", size = 0.5) 
  ) +
  coord_cartesian(xlim = c(200, 8000)) 
```

Again, this plot shows the average absorbance over frequencies ranging from 200 to 8000. The 2 lines in the plot represent Male and Female tested subjects. As shown by the plot, there is almost no difference in the mean absorbance rates based on gender.
