---
title: "WAI Database Study"
description: |
  Analysis of Sound Absorbance Data from WAI Database
author: Paityn Richardson
date: November 26, 2024
format: html
execute: 
  warning: false
  message: false
---

```{r}
library(RMariaDB)
library(tidyverse)
con_wai <- dbConnect(
  MariaDB(), host = "scidb.smith.edu",
  user = "waiuser", password = "smith_waiDB", 
  dbname = "wai"
)
Measurements <- tbl(con_wai, "Measurements")
PI_Info <- tbl(con_wai, "PI_Info")
Subjects <- tbl(con_wai, "Subjects")

# collect(Measurements)
```

In this project I plan to use the data from the [Wideband Acoustic Immittance (WAI) Database](https://www.science.smith.edu/wai-database/) hosted by Smith College. I plan to replicate a figure published in the Official Journal of the American Auditory Society by Susan Voss titled "Mean Absorbance from each Publication in the WAI Database" [Voss(2020)](https://journals.lww.com/ear-hearing/citation/2019/11000/resource_review.21.aspx). The database contains 3 data sets: Measurements, PI_Info, and Subjects, which all have different information pertaining to the study.

Here is the measurements table,

```{sql}
#| connection: con_wai
DESCRIBE Measurements
```

Here is the PI_Info table,

```{sql}
#| connection: con_wai
DESCRIBE PI_Info
```

And here is the Subjects table,

```{sql}
#| connection: con_wai
DESCRIBE Subjects
```

I started of by arranging the data using a SQL query.

```{sql}
#| connection: con_wai
#| output.var: "new_table"
    SELECT 
        m.Frequency,
        m.Identifier,
        m.Instrument,
        AVG(m.Absorbance) AS Mean_Absorption, 
        CONCAT(pi.AuthorsShortList, ' (', pi.Year, ')', ' N=', COUNT(DISTINCT CONCAT(m.SubjectNumber, '-', m.Ear)),'; ', m.Instrument) AS Label
    FROM 
        Measurements m
    JOIN 
        PI_Info pi
        ON m.Identifier = pi.Identifier
    WHERE 
        m.Identifier IN ('Abur_2014', 'Feeney_2017', 'Groon_2015', 'Lewis_2015', 'Lui_2008', 'Rosowski_2012', 'Shahnaz_2006', 'Shaver_2013', 'Sun_2016', 'Voss_1994', 'Voss_2010', 'Werner_2010') 
        AND m.Frequency BETWEEN 200 AND 8000
    GROUP BY 
        m.Frequency, 
        m.Identifier, 
        m.Instrument
    ORDER BY 
        m.Frequency, 
        m.Identifier, 
        m.Instrument


```

In this query, I grouped the data by Frequency, Identifier, and Instrument. In order to get the variable "AuthorShortList" I joined the PI_Info table and the Measurements table on SubjectNumber. I used Concat to make a label for each of the studies with the number of unique ears, the instrument used, and the year and author of the study. My goal was to match the figure created by Voss in 2022, so I filtered for only studies included in that specific plot.

```{r}
library(ggplot2)


ggplot(new_table, aes(x = Frequency, y = Mean_Absorption, color = Label)) +
  geom_line(size = .7, na.rm = TRUE) +  
  scale_x_log10(name = "Frequency (Hz)",
    breaks = c(200, 400, 600, 800, 1000, 2000, 4000, 6000, 8000),
    labels = c("200", "400", "600", "800", "1000", "2000", "4000", "6000", "8000")) +
  scale_y_continuous(
    name = "Mean Absorbance",
    limits = c(0, 1), 
    breaks = seq(0, 1, by = 0.1)  
  ) +
  labs(
    title = "Mean Absorbance From Each Publication in the WAI Database",  
    color = NULL 
  ) +
  theme_minimal() + 
  theme(
    axis.text = element_text(size = 5),
    axis.title = element_text(size = 7),
    plot.title = element_text(size = 9, face = "bold", hjust = .5),
    legend.text = element_text(size = 5),
    legend.position = c(0.28, 0.9),
    legend.key.size = unit(0.1, "cm"),
    legend.spacing = unit(0.01, "cm"),
    legend.background = element_rect(color = "black", size = .2, fill = "white"),
    aspect.ratio = 1
  ) 



```

This table shows the mean absorbance of frequencies ranging from 200 HZ to 8000 HZ in 12 different studies in the WAI database. As seen in the table, absorbance, on average, increases as frequency increases up until around 1000 HZ, when it drops off and begins to decrease again.

```{sql}
#| connection: con_wai
#| output.var: "new_table4"
SELECT 
    m.Frequency, 
    s.Sex, 
    AVG(m.Absorbance) AS Avg_Absorbance
FROM 
    Measurements m
JOIN 
    Subjects s
ON 
    m.SubjectNumber = s.SubjectNumber
WHERE 
    m.Identifier = 'Aithal_2015'
    AND s.Sex != 'Unknown'
    AND m.Frequency BETWEEN 200 AND 8000
GROUP BY 
    m.Frequency, s.Sex
ORDER BY 
    m.Frequency, s.Sex;

```

Now with this SQL query, I am doing a similar thing to the last, except now I am also grouping by "sex" which required me to join the table "Subjects" also in the WAI database. For this table i chose to focus on 1 study in the WAI data base, this being [A Comparison With High-Frequency Tympanometry, Automated Brainstem Response, and Transient Evoked and Distortion Product Otoacoustic Emissions](https://journals.lww.com/ear-hearing/abstract/2015/09000/wideband_absorbance_outcomes_in_newborns__a.18.aspx)(Aithal 2015).

```{r}
dbDisconnect(con_wai, shutdown = TRUE)
```

```{r}
library(ggplot2)

ggplot(new_table4, aes(x = Frequency, y = Avg_Absorbance, color = Sex)) +
  geom_line(size = 0.7, na.rm = TRUE) +
  scale_x_log10(name = "Frequency (Hz)",
    breaks = c(200, 400, 600, 800, 1000, 2000, 4000, 6000, 8000),
    labels = c("200", "400", "600", "800", "1000", "2000", "4000", "6000", "8000")) +
  scale_y_continuous(
    name = "Average Absorbance",
    limits = c(0, 1), 
    breaks = seq(0, 1, by = 0.1)
  ) +
  labs(
    title = "Average Absorbance by Frequency and Gender(Aithal 2015)",
    color = NULL 
  ) +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 8),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 8),
    legend.position = c(0.1, 0.9),
    legend.box.background = element_rect(color = "black", fill = "white", size = 0.5) 
  ) +
  coord_cartesian(xlim = c(200, 8000)) 
```

Again, this plot shows the average absorbance over frequencies ranging from 200 to 8000. The 2 lines in the plot represent Male and Female tested subjects. As shown by the plot, there is almost no difference in the mean absorbance rates based on gender.
